Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-01-27 ms.Rnw
changes to genealogy details

\subsubsection{Jump marks}

In the following, it will be useful to break the jumps into distinct categories.
For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
\begin{equation*}
  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
\end{equation*}
In \cref{fig:markov_state}, we use the marks to distinguish events with distinct effects on genealogical relationships.
Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.

Let us define the \emph{jump mark} process, $\Ur_t$, to be the mark of the latest jump as of time $t$.
As usual, we take the sample paths of $\Ur_t$ to be \cadlag.
Observe that $(\Xr_t,\Ur_t)$ is a Markov process, though $\Ur_t$ is not.

@@

\subsection{History process}

@@

For our purposes, a \emph{genealogy}, $\G_{T}$, is a labeled, time-calibrated tree.
It encapsulates the relationships of shared ancestry among a set of lineages that are alive at time $T\in\Rp$, and perhaps a set of samples collected at earlier times.
There are three distinct kinds of nodes:
\begin{inparaenum}[(i)]
\item \emph{tip nodes}, which represent labeled living lineages;
\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another; and
\item \emph{sample nodes}, which represent labeled samples.
\end{inparaenum}
Each node $a$ is associated with a specific time, $t(a)$.
In particular, if $a$ is a tip node in $\G_T$, then $t(a)=T$;
if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
Moreover, if node $a$ is ancestral to node $a'$, then $t(a)\le{t(a')}$ and $t(a')-t(a)$ is the distance between $a$ and $a'$ along the genealogy.
We let $\event{\G}{T}$ denote the set of all node-times of the genealogy $\G_T$.
Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme.

Formally, we define a genealogy, $\G_T$, to be an ordered pair $(y,z)$.
The $z$ element specifies the topological structure, as follows.
Let $\leaves{\G_T}$ be the set of all tip and sample nodes in $\G_T$ and define $\func{P}$ to be the set of all partitions of $\leaves{\G_T}$.
The partition \emph{fineness} defines a partial order on $\func{P}$.
Specifically, for $p,p'\in{\func{P}}$, we say $p\preceq{p'}$ if and only if for every $c'\in{p'}$ there is a unique $c\in{p}$ such that $c\supseteq{c'}$.
The topological structure of $\G_T$ is defined by a map $z:[0,T]\to\func{P}$ that is \cadlag\ and monotone in the sense that $t_1<t_2$ implies $\z_{t_1}\preceq{\z_{t_2}}$.

The $y$ element of $\G_T$ specifies the coloring of branches and location of nodes.
In particular, for $a\in\leaves{\G_T}$ and $t\in[0,T]$, define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$.
To make this well-defined, let $d_{t}(a)=\eth$ for $t\in{[t(a),T]}$;
we call $\eth$ the \emph{underdeme} and define $\Demesplus\coloneq{\Demes\cup\{\eth\}}$.
Thus $d:[0,T]\times\leaves{\G_T}\to\Demesplus$ is a well-defined \cadlag\ function.

Since not every node along a lineage corresponds to a change of deme, we also count the number of nodes along each lineage.
Let $m:[0,T]\times\leaves{\G_T}\to\Zp$ be such that $m_{t}(a)$ is the number of internal or sample nodes encountered along lineage $a$ as one goes from time $0$ to time $t\le{T}$ along the genealogy.
For $t\in{[t(a),T]}$, let $m_{t}(a)=m_{t(a)}(a)$.
Clearly, $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.

Finally, we define the map $y:[0,T]\times\leaves{\G_T}\to\Demesplus\times\Zp$ by $y_{t}(a)=\left(d_{t}(a),m_{t}(a)\right)$.
The two \cadlag\ functions $y$ and $z$ specify the genealogy $\G_T$ completely and uniquely.
When the genealogy is not clear from context, we write $y^{\G}$ and $z^{\G}$ to refer to the respective components of genealogy $\G$.

Clearly, the topological structure of a tree imposes constraints on the permissible functions $y$ that describe the painting and node placement along it.
In particular, $y$ must respect the topological structure of $\G_T$, such that if two lineages share a common ancestor at time $t$, both lineages must have been in the same deme at that time and the number of previous events along the lineage must agree.
In other words, if $a,b\in{p}\in{\z_{t}}$, then $y_{t}(a)=y_{t}(b)$.
Let $\Yspace^z$ denote the space of all functions that obey this constraint at every $t$.
Let $\Yspace^\z_t\subseteq{\Demesplus\times\Zp}$ be the time-$t$ \emph{section} of $\Yspace^z$, \ie $\Yspace^\z_t=\displaystyle\bigcup_{y\in\Yspace^z}{\Set{y_t}}$.

\subsection{Jump types}
\label{sec:jump_types}

We will show how a given population process naturally induces a process in the space of genealogies.
Because different kinds of jumps have different effects on the genealogy, we need a way of distinguishing them.
In particular, there are five distinct \emph{pure types} of jumps:
\begin{compactenum}[(a)]
\item \emph{Birth-type jumps} result in the branching of one or more new lineages, each from some existing lineages.
  Examples of birth-type events include transmission events, speciations, and actual births.
  For simplicity, we assume that all new lineages arising from a birth event share the same parent.
\item \emph{Death-type jumps} result in the extinction of one or more lineages.
  Examples include recovery from infection, death of a host, and species extinctions.
  We allow for the possibility that multiple lineages die simultaneously.
\item \emph{Migration-type jumps} result in the movement of a lineage from one deme to another.
  Spatial movements, changes in host age or behavior, and progression of an infection can all be represented as migration-type events.
  We permit multiple movements to occur simultaneously.
\item \emph{Sample-type jumps} result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
  We allow for the possibility that multiple samples are collected in a single sample event.
  In this case, we assume that, in each sample-type event, each extant lineage is sampled at most once.
\item \emph{Neutral-type jumps} result in no change to any of the lineages.
\end{compactenum}
\Cref{fig:markov_state} depicts an example with all five of the pure types.
It is not necessary that a jump be of a pure type.
Compound jumps partake of more than one type.
For example, a sample/death-type event, in which a lineage is simultaneously sampled and removed, has been proposed \citep{Leventhal2014}, as have birth/death events in which one lineage reproduces at the same moment that another dies (\eg the Moran process \citep{Moran1958}).
The theory presented here places few restrictions on the complexity of the events that can occur.

@@

We now show how a given population process induces a stochastic process, $\Gen_t$, on the space of genealogies.
In the case of unstructured population processes (\ie those having a single deme), \citet{King2022} gave a construction.
Although we now treat a more general case, the ideas are much the same.
Readers wishing more detail on the construction here should consult the earlier paper.

@@

At each event in the population process, one or more of the following changes happen to the genealogy, according to the type of the event (\cref{fig:event_types}):
\begin{compactenum}[(a)]
\item
  A birth-type event at time $t$ (\cref{fig:event_types}A) results in the creation of one new internal node, call it $b$.
  A tip node, $a$, of the appropriate deme is chosen with uniform probability from among those present and $b$ is inserted so that its ancestor is that of $a$, while $a$ takes $b$ as its ancestor.
  One new tip node, of the appropriate deme, is created for each of the children, all of which take $b$ as their immediate ancestor.
\item
  In a death-type event (\cref{fig:event_types}B), one or more tip nodes of the appropriate demes are selected with uniform probability from among those present.
  These are deleted.
  Next, branch nodes without children are recursively removed.
  Sample nodes are never removed.
\item
  In a migration-type event (\cref{fig:event_types}C), the appropriate number of migrating lineages are selected at random, with uniform probability according to deme, from among those present.
  For each selected lineage, one new branch node is inserted between the selected tip node and its ancestor.
  The color of the descendant branch changes accordingly.
\item
  At a sample-type event (\cref{fig:event_types}D), the appropriate number of sampled lineages are selected at random from among the tip nodes, with uniform probability according to deme.
  One new sample node is introduced for each selected lineage:
  each is inserted between a selected tip nodes and its ancestor.
\item
  At a neutral-type event (\cref{fig:event_types}E), no change is made to the genealogy.
\item
  Finally, events of compound type (\eg \cref{fig:event_types}F--H) are accommodated by combining the foregoing rules.
\end{compactenum}
In each of these events, the new node or nodes that are introduced are situated at time $t$.

@@

The lineages which descend from an inserted node are said to \emph{emerge} from the event.
Thus, after a birth-type event, the emerging lineages include all the new offspring as well as the parent.
Likewise, at pure migration- or sample-type events, each migrating or sampled lineage emerges from the event.
At pure death-type events, no lineages emerge.
In general, at an event of mark $u$, there are $r^u_i$ emergent lineages in deme $i$.
We require that $r^u_i$ be a constant, for each $u$ and $i$.
Since one is free to expand the set of jump-marks $\Jumps$ as needed, this is not an important restriction on the models that the theory can accommodate.
Thus there is a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
We say $r^u=(r^u_i)_{i\in\Demes}$ is the \emph{production} of an event of mark $u$.
Note that the lineages that die as a result of an event do not count in the production but that all parent lineages, if they survive the event, are counted in the production.

@@

Given a genealogy $\Gen_t$, one obtains the \emph{pruned genealogy}, $\Pr_t=\prune(\Gen_t)$ by first dropping every tip node and then recursively dropping every childless internal node.
In a pruned genealogy only internal and sample nodes remain, and sample nodes are found at all of the leaves and possibly some of the interior nodes of the genealogy.
Observe that a pruned genealogy is a colored genealogy:
it retains information about where among the demes each of its lineages was through time.
Note also that a pruned genealogy $\Pr_t$ is characterized by functions $y^{\Pr_t}$ and $z^{\Pr_t}$ just as an un-pruned genealogy is.

@@

In the following, we will find that we need to count the deme-specific numbers of lineages present in a given pruned genealogy at a given time.
Accordingly, suppose $\Pr_T=(y,z)$ is a pruned genealogy and suppose $t\in[0,T]$.
Let $\ell_i$ denote the number of lineages in deme $i$ at time $t$ and $\ell=(\ell_i)_{i\in\Demes}\in\Zp^{\Demes}$.
Clearly, $\ell$ depends only on the values, $(y_t,\z_t)$, of $y$ and $z$ at time $t$.
Therefore, we can define $\ell$ as a function such that, whenever $\Pr_T=(y,z)$ is a pruned genealogy, $\ell(y_t,\z_t)$ is the vector of deme-specific lineage counts at time $t$.
We refer to $\ell$ as the \emph{lineage-count} function (cf.~\cref{fig:ells}).

We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
In particular, given a node time $t$ in a pruned genealogy $\Pr_T=(y,z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
Like the lineage-count function, $s$ depends only on the local structure of $\Pr_T$.
However, $s$ depends not only on $(\y_t,\z_t)$, but also on $(\yt_t,\zt_t)$.
Thus, we can define the \emph{saturation} function such that, whenever $\Pr_T=(\y,\z)$ is a pruned genealogy, $s(\yt_t,\y_t,\zt_t,\z_t)$ is the integer vector of deme-specific numbers of emerging lineages at time $t$.
\Cref{fig:ells} illustrates.

@@

Suppose $\P_T$ is a pruned genealogy and $t\in\event{P}{T}$.
The local structure of $\P_T$ is, in general, compatible with only a subset of the possible jumps $\Jumps$.
For example, if the event in $\P_T$ is a branch node or a sample node, then it is compatible only with birth-type or sample-type jumps, respectively.
Similarly, if the node in $\P_T$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
To accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $y$ and $z$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
That is, $Q_u(y,y',z,z')=1$ if and only if
there is a feasible genealogy, $\G_T$, and history, $\H_T$,
and a $t\in{[0,T]}$ such that,
given $\Gr_T=\G_T$ and $\Hr_T=\H_T$,
$\U_t=u$
, $\yt^{\prune(\G_T)}_t=y$, $y^{\prune(\G_T)}_t=y'$,
$\zt^{\prune(\G_T)}_t=z$, and $z^{\prune(\G_T)}_t=z'$.
We refer to $Q$ as the \emph{compatibility indicator}.

@@

The \emph{obscured genealogy} is obtained by discarding all information about demes and events not visible from the topology of the tree alone.
In particular, if $\Pr_T=(y,z)$ is a pruned genealogy, we write $\obs(\Pr_T)=z$ to denote the obscured genealogy.

@@

\begin{thm}\label{thm:obsc_lik}
  Let $\z^*=\obs(\P^*_T)$ be a given obscured genealogy.
  Let $\pi$ and $q$ be probability kernels, such that for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $\pi_u(t,x,x',y,y')\ge{0}$, $\sum_{y'\in\Yspace_t^{z^*}}{\pi_u(t,x,x',y,y')}=1$, $q(x,y)\ge{0}$, and $\sum_{y\in\Yspace_0^{\z^*}}{q(x,y)}=1$.
  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $Q_u(y,y',\zt^*_t,\z^*_t)=1$
  and that $q(x,y)>0$ whenever $\Prob{\Xr_0=x,\y^{\Pr_T}_0=y}>0$.
  Then there is a Markov jump process $\yr_t$ on $\Yspace_{\z^*}$ such that
  \begin{equation*}
    \CondProb{\zr_T=\z^*}{\Hr_T=\H_T}=
    \Indicator{\event{\H}{T}\supseteq\event{\z}{T}^*}\,\Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{T}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}},
  \end{equation*}
  where $\phi$ is defined in \cref{eq:phidef} and
  the expectation is taken over the sample paths of $\yr_t$, $t\in[0,T]$.
\end{thm}
\begin{proof}
  First, observe that, since $\obs$ is a deterministic operator,
  \begin{equation}\label{eq:IS1}
    \CondProb{\zr_T=z^*}{\Hr_T}=\CondExpect{\Indicator{\obs(\Pr_T)=z^*}}{\Hr_T}.
  \end{equation}
  Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
  we will propose pruned genealogies compatible with $z^*_t$ as sample paths from a Markov process on $\Yspace_{z^*}$ and
  evaluate the the expectation in \cref{eq:IS1} by summing over these paths.
  Conditional on $\Hr_T$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$, on $\Yspace_{z^*}$ such that
  \begin{equation*}
    \begin{gathered}
      \Prob{\yrh_0}=q(\yrh_0),
      \qquad
      \CondProb{\yrh_k}{\yrh_{k-1},\Hr_T=\H_T}=\pi_{\Uh_k}(\Th_k,\Xh_{k-1},\Xh_{k},\yrh_{k-1},\yrh_k).
    \end{gathered}
  \end{equation*}
  The required Markov process $\yr_t$ is the unique \cadlag\ process with event times $\Trh_k$ and $\yrh_k$ as its embedded chain.
  This construction of $\yr_t$ obviously guarantees that $\event{\H}{t}\supseteq\event{\y}{}\supseteq\event{\z}{}^*$.

>>>>

\subsection{Jump marks}

In the following, it will be useful to break the jumps of the population process $\Xr_t$ into distinct categories.
For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
\begin{equation*}
  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
\end{equation*}
In \cref{fig:markov_state}, we use the marks to distinguish events with distinct effects on genealogical relationships.
Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.

Let us define the \emph{jump mark} process, $\Ur_t$, to be the mark of the latest jump as of time $t$.
As usual, we take the sample paths of $\Ur_t$ to be \cadlag.
Observe that, though $\Xr_t$ and $(\Xr_t,\Ur_t)$ are Markov processes, $\Ur_t$ is not.

\subsection{History}

\AAK{[Maybe it would be better to speak instead of $\Hr_t$ being the $\sigma$-algebra generated by the sequences $(\Trh_k,\Xrh_k,\Urh_k)_{k=1}^{\Kr_t}$.
    The only thing we do with $\Hr$ is condition on it.
    One (perhaps minor) problem with this is that we cannot then speak about the measure on histories.
]}

@@

@@

For our purposes, a \emph{genealogy}, $\G_{T}$, is a labeled, time-calibrated tree.
It encapsulates the relationships of shared ancestry among a set of lineages that are alive at time $T\in\Rp$, and perhaps a set of samples collected at earlier times.
A genealogy includes three distinct kinds of nodes:
\begin{inparaenum}[(i)]
\item \emph{tip nodes}, which represent labeled living lineages;
\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another; and
\item \emph{sample nodes}, which represent labeled samples.
\end{inparaenum}
Each node $a$ is associated with a specific time, $t(a)$.
In particular, if $a$ is a tip node in $\G_T$, then $t(a)=T$;
if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
Moreover, if node $a$ is ancestral to node $a'$, then $t(a)\le{t(a')}$ and $t(a')-t(a)$ is the distance between $a$ and $a'$ along the genealogy.
We let $\event{\G}{T}$ denote the set of all node-times of the genealogy $\G_T$.
Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme.

Formally, we define a genealogy, $\G_T$, to be an ordered pair $(\y,\z)$.
The $\z$ element specifies the genealogy's tree structure, as follows.
Let $\leaves$ be a countable set of labels and let $\func{Z}$ be the set of all partitions of subsets of $\leaves$.
Partition \emph{fineness} defines a partial order on $\func{Z}$.
Specifically, for $\zeta,\zeta'\in{\func{Z}}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
The tree structure of $\G_T$ is defined by a \cadlag\ map $\z:[0,T]\to\func{Z}$ that is monotone in the sense that $t_1\le{t_2}$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
An element $b\in\z_t$ is a set of labels;
it represents the branch of the tree that bears the corresponding lineages.

The $\y$ element of $\G_T$ specifies the coloring of branches and location of nodes.
In particular, if $a$ is the label of any tip or sample node and $t\in[0,T]$, we define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$ and think of this as the color of the corresponding branch at that time.
We will also need to count the number of nodes along each lineage:
let $m_{t}(a)$ be the number of internal or sample nodes encountered as one descends along lineage $a$ from time $t=t(a)$ to $t=0$.
Observe that $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.
Formally, we define the coloring component $\y$ as follows.
Let $\Yspace=\Demes\times\Zp$.
For $\eta\in\Yspace$, write $\eta=(\eta^d,\eta^m)$ to denote the two components.
Then, for $\zeta\in\func{Z}$, each $\eta\in\Yspace^{\zeta}$ is a map from $\zeta$ into $\Yspace$.
Now, given a tree $\z:[0,T]\to\func{Z}$ and a map $\y:[0,T]\to{\coprod_{t\in[0,T]}{\Yspace^{\z_t}}}$,
we say that $\y$ is \emph{compatible} with $\z$ if,
for every $t\in[0,T]$, $\y_t\in{\Yspace^{\z_t}}$,
$\y$ is \cadlag\ and,
whenever $t_1,t_2\in[0,T]$, $b_1\in\z_{t_1}$, $b_2\in\z_{t_2}$
such that $t_1\le{t_2}$ and $b_1\supseteq{b_2}$,
we have $y^m_{t_1}\le{y^m_{t_2}}$.
Let $\func{Y}(\z)$ denote the set of such maps compatible with $\z$.
\AAK{[The coproduct is a fiber bundle over the space of trees and $\func{Y}(\z)$ is a section of this bundle.]}

The two \cadlag\ functions $\y$ and $\z$ specify a genealogy completely and uniquely.
When the genealogy is not clear from context, we write $\G^{\y}$ and $\G^{\z}$ to refer to the respective components of genealogy $\G$.

\subsection{Event types}
\label{sec:event_types}

We will show how a given population process naturally induces a process in the space of genealogies.
Specifically, we will describe how, at each jump in the population process, a corresponding change occurs in the genealogy, according to lineages branch, die, or move between demes and whether samples are taken.
For this purpose, there are five distinct \emph{pure types} of events:
\begin{compactenum}[(a)]
\item \emph{Birth-type events} result in the branching of one or more new lineages, each from some existing lineages.
  Examples of birth-type events include transmission events, speciations, and actual births.
  For simplicity, we assume that all new lineages arising from a birth event share the same parent.
\item \emph{Death-type events} result in the extinction of one or more lineages.
  Examples include recovery from infection, death of a host, and species extinctions.
  We allow for the possibility that multiple lineages die simultaneously.
\item \emph{Migration-type events} result in the movement of a lineage from one deme to another.
  Spatial movements, changes in host age or behavior, and progression of an infection can all be represented as migration-type events.
  We permit multiple movements to occur simultaneously.
\item \emph{Sample-type events} result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
  We allow for the possibility that multiple samples are collected in a single sample event.
  In this case, we assume that, in each sample-type event, each extant lineage is sampled at most once.
\item \emph{Neutral-type events} result in no change to any of the lineages.
\end{compactenum}
\Cref{fig:markov_state} depicts an example with jumps of all five pure types.
It is not necessary that an event be of a pure type;
\emph{compound events} partake of more than one type.
For example, a sample/death-type event, in which a lineage is simultaneously sampled and removed, has been proposed \citep{Leventhal2014}, as have birth/death events in which one lineage reproduces at the same moment that another dies (\eg the \citet{Moran1958} process).
The theory presented here places few restrictions on the complexity of the events that can occur by combining events of the various pure types.

@@

We now show how a given population process induces a stochastic process, $\Gen_t$, on the space of genealogies.
In the case of unstructured population processes (\ie those having a single deme), \citet{King2022} gave a related construction that is equivalent to the one presented here.

@@

At each jump in the population process, a change is made to the genealogy, according to the mark, $u$, of the jump (\cref{fig:event_types}).
In particular:
\begin{compactenum}[(a)]
\item
  If $u$ is of birth-type (\cref{fig:event_types}A), it results in the creation of one new internal node, call it $b$.
  A tip node, $a$, of the appropriate deme is chosen with uniform probability from among those present and $b$ is inserted so that its ancestor is that of $a$, while $a$ takes $b$ as its ancestor.
  One new tip node, of the appropriate deme, is created for each of the children, all of which take $b$ as their immediate ancestor.
\item
  If $u$ is of death-type (\cref{fig:event_types}B), one or more tip nodes of the appropriate demes are selected with uniform probability from among those present.
  These are deleted.
  Next, branch nodes without children are recursively removed.
  Sample nodes are never removed.
\item
  At a migration-type event (\cref{fig:event_types}C), the appropriate number of migrating lineages are selected at random with uniform probability, from among those present in the appropriate demes.
  For each selected lineage, one new branch node is inserted between the selected tip node and its ancestor.
  The color of the descendant branch changes accordingly.
\item
  At a sample-type event (\cref{fig:event_types}D), the appropriate number of sampled lineages are selected at random from among the tip nodes, with uniform probability according to deme.
  One new sample node is introduced for each selected lineage:
  each is inserted between a selected tip nodes and its ancestor.
\item
  At a neutral-type event (\cref{fig:event_types}E), no change is made to the genealogy.
\item
  Finally, events of compound type (\eg \cref{fig:event_types}F--H) are accommodated by combining the foregoing rules.
\end{compactenum}
In each of these events, the new node or nodes that are introduced have node-times equal to the time of the jump.

@@

The lineages which descend from an inserted node are said to \emph{emerge} from the event.
Thus, after a birth-type event, the emerging lineages include all the new offspring as well as the parent.
Likewise, at pure migration- or sample-type events, each migrating or sampled lineage emerges from the event.
At pure death-type events, no lineages emerge.
In general, at an event of mark $u$, there are $r^u_i$ emergent lineages in deme $i$.
We require that $r^u_i$ be a constant, for each $u$ and $i$.
Since, in applications, one is free to expand the set of jump-marks $\Jumps$ as needed, this is not an important restriction on the models that the theory can accommodate.
Thus there is a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
We say $r^u=(r^u_i)_{i\in\Demes}$ is the \emph{production} of an event of mark $u$.
Note that the lineages that die as a result of an event do not count in the production but that all parent lineages, if they survive the event, are counted in the production.

@@

Given a genealogy $\Gen_t$, one obtains the \emph{pruned genealogy}, $\Pr_t=\prune(\Gen_t)$ by first dropping every tip node and then recursively dropping every childless internal node.
In a pruned genealogy only internal and sample nodes remain, and sample nodes are found at all of the leaves and possibly some of the interior nodes of the genealogy.
Observe that a pruned genealogy is a colored genealogy:
it retains information about where among the demes each of its lineages was through time.
Note also that a pruned genealogy $\Pr_t$ is characterized by functions $\Pr_t^\y$ and $\Pr_t^\z$ just as an un-pruned genealogy is.

@@

In the following, we will find that we need to count the deme-specific numbers of lineages present in a given pruned genealogy at a given time.
Accordingly, suppose $\Pr_T=(y,z)$ is a pruned genealogy and suppose $t\in[0,T]$.
Let $\ell_i$ denote the number of lineages in deme $i$ at time $t$ and $\ell=(\ell_i)_{i\in\Demes}\in\Zp^{\Demes}$.
Clearly, $\ell$ depends only on the values, $(\y_t,\z_t)$, of $y$ and $z$ at time $t$.
Therefore, we can define $\ell$ as a function such that, whenever $\Pr_T=(\y,\z)$ is a pruned genealogy, $\ell(\y_t,\z_t)$ is the vector of deme-specific lineage counts at time $t$.
We refer to $\ell$ as the \emph{lineage-count} function (cf.~\cref{fig:ells}).

We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
In particular, given a node time $t$ in a pruned genealogy $\Pr_T=(\y,\z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
Like the lineage-count function, $s$ depends only on the local structure of $\Pr_T$.
However, $s$ depends not only on $(\y_t,\z_t)$, but also on $(\yt_t,\zt_t)$.
Thus, we can define the \emph{saturation} function such that, whenever $\Pr_T=(\y,\z)$ is a pruned genealogy, $s(\yt_t,\y_t,\zt_t,\z_t)$ is the integer vector of deme-specific numbers of emerging lineages at time $t$.
\Cref{fig:ells} illustrates.

@@

Suppose $\P_T$ is a pruned genealogy and $t\in\event{P}{T}$.
The local structure of $\P_T$ is, in general, compatible with only a subset of the possible jumps $\Jumps$.
For example, if the event in $\P_T$ is a branch node or a sample node, then it is compatible only with birth-type or sample-type jumps, respectively.
Similarly, if the node in $\P_T$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
To accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $y$ and $z$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
That is, $Q_u(\eta,\eta',\zeta,\zeta')=1$ if and only if
there is a feasible genealogy, $\G_T=(\y,\z)$, and history, $\H_T$,
and a $t\in{[0,T]}$ such that,
given $\Gr_T=\G_T$ and $\Hr_T=\H_T$,
$\U_t=u$,
$\yt_t=\eta$, $\y_t=\eta'$,
$\zt_t=\zeta$, and $\z_t=\zeta'$.
We refer to $Q$ as the \emph{compatibility indicator}.

@@

The \emph{obscured genealogy} is obtained by discarding all information about demes and events not visible from the topology of the tree alone.
In particular, if $\Pr_T=(\y,\z)$ is a pruned genealogy, we write $\obs(\Pr_T)=\Pr_T^{\z}=\z$ to denote the obscured genealogy.

@@

\begin{thm}\label{thm:obsc_lik}
  \AAK{[This can be rephrased somewhat more compactly into a result claiming the existence of the Markov process $\y$.]}
  Let $\z^*=\obs(\P^*_T)$ be a given obscured genealogy.
  Let $\pi$ and $q$ be probability kernels, such that for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $\pi_u(t,x,x',y,y')\ge{0}$, $\sum_{y'\in\Yspace_t^{z^*}}{\pi_u(t,x,x',y,y')}=1$, $q(x,y)\ge{0}$, and $\sum_{y\in\Yspace_0^{\z^*}}{q(x,y)}=1$.
  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $Q_u(y,y',\zt^*_t,\z^*_t)=1$
  and that $q(x,y)>0$ whenever $\Prob{\Xr_0=x,\y^{\Pr_T}_0=y}>0$.
  Then there is a Markov jump process $\yr_t$ on $\Yspace_{\z^*}$ such that
  \begin{equation*}
    \CondProb{\zr_T=\z^*}{\Hr_T=\H_T}=
    \Indicator{\event{\H}{T}\supseteq\event{\z}{T}^*}\,\Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{T}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}},
  \end{equation*}
  where $\phi$ is defined in \cref{eq:phidef} and
  the expectation is taken over the sample paths of $\yr_t$, $t\in[0,T]$.
\end{thm}
\begin{proof}
  First, observe that, since $\obs$ is a deterministic operator,
  \begin{equation}\label{eq:IS1}
    \CondProb{\zr_T=z^*}{\Hr_T}=\CondExpect{\Indicator{\obs(\Pr_T)=z^*}}{\Hr_T}.
  \end{equation}
  Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
  we will propose pruned genealogies compatible with $z^*_t$ as sample paths from a Markov process on $\Yspace_{z^*}$ and
  evaluate the the expectation in \cref{eq:IS1} by summing over these paths.
  Conditional on $\Hr_T$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$, on $\Yspace_{z^*}$ such that
  \begin{equation*}
    \begin{gathered}
      \Prob{\yrh_0}=q(\yrh_0),
      \qquad
      \CondProb{\yrh_k}{\yrh_{k-1},\Hr_T=\H_T}=\pi_{\Uh_k}(\Th_k,\Xh_{k-1},\Xh_{k},\yrh_{k-1},\yrh_k).
    \end{gathered}
  \end{equation*}
  The required Markov process $\yr_t$ is the unique \cadlag\ process with event times $\Trh_k$ and $\yrh_k$ as its embedded chain.
  This construction of $\yr_t$ obviously guarantees that $\event{\H}{t}\supseteq\event{\y}{}\supseteq\event{\z}{}^*$.
